---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# IRLSpoly

The traditional algorithms for estimating the polyserial/polychoric correlation coefficients are designed based on the maximum likelihood method. Although they are accurate, they are very slow due to the numerical calculation of double integral in iteration.

IRLS provides an iterative regression algorithm based on conditional mathematical expectation, which can calculate the polyserial/polychoric correlation coefficient extremely quickly with similar accuracy.

The paper explaining the theory is placed in \url{https://arxiv.org/abs/2210.11115}.


## Installation

You can install the development version of IRLSpoly from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("encoreus/IRLSpoly")
```

```{r}
options (warn = -1)
library(IRLSpoly)
library(psych)
```

## Simmulation

You can use gen_polyserial and gen_polychoric to generate simulated polyserial/polychoric sequences:
```{r}
set.seed(1997)
X1 = gen_polyseries(500,0.5,-1:1)
X2 = gen_polychoric(500,0.5,-1:0,0:1)
X1[,1:6]
X2[,1:6]
table(X2[1,],X2[2,])
```

You can use esti_polyserial and esti_polychoric to estimate polyserial/polychoric coefficients:
```{r}
result1 = esti_polyserial(X1)
result2 = esti_polychoric(X2)

print("The rhohat of polyserial coefficient is:")
print(result1$rho)
print("The standard error of polyserial coefficient is:")
print(result1$std)

print("The rhohat of polyserial coefficient is:")
print(result2$rho)
print("The standard error of polyserial coefficient is:")
print(result2$std)
```

Compare with the MLE in package: polycor:
```{r}
MLE1 = polycor::polyserial(X1[1,],X1[2,],ML=T,std.err=T)
MLE2 = polycor::polychor(X2[1,],X2[2,],ML=T,std.err=T)

print("The rhohat of polyserial coefficient is:")
print(MLE1$rho)
print("The standard error of polyserial coefficient is:")
print(sqrt(MLE1$var[1,1]))

print("The rhohat of polyserial coefficient is:")
print(MLE2$rho)
print("The standard error of polyserial coefficient is:")
print(sqrt(MLE2$var[1,1]))
```

The outcomes of IRLS provide support points sequence during iterative regression. Draw these points to visualize the convergence process:
```{r}
set.seed(1998)
x1<-X2[1,]
x2<-X2[2,]
r0<-cor(x1,x2)
rho=0.5

# support points
EX <-result2$Ex
EY = result2$Ey
iter = result2$iter
lm1 <- lm(EY[,iter]~0 + EX[,iter])
plot(x1,x2,pch=19,xlim=c(-1.5,1.5),ylim=c(-1.5,1.5),xlab=expression(Z[1]),ylab=expression(Z[2]),col='gray')
points(cbind(EX[1,1:iter],EY[1,1:iter]),pch=3,col='red',type='b',lty=3)
points(cbind(EX[2,1:iter],EY[2,1:iter]),pch=3,col='blue',type='b',lty=3)
points(cbind(EX[3,1:iter],EY[3,1:iter]),pch=3,col='purple',type='b',lty=3)

# threshold
abline(a=0,b=r0,col='red')
abline(a=0,b=rho,col='green')
abline(lm1)
abline(h=0)
abline(h=1)
abline(v=-1)
abline(v=0)
legend("topleft", inset=0.03, text.width=0.5, x.intersp=0.3, c("pearson correlation","true correlation", "polychoric correlation"),
               lty=c(1, 1, 1), cex=0.8,col=c("red", "green", 'black'))
grid()
```

## Data Analysis

The first two dataset is from package psych:

25 Personality items representing 5 factors
```{r}
bfi.com <- bfi[complete.cases(bfi[1:5]),1:5]
print("Pearson:")
lowerCor(bfi.com) #the default Pearson
print("Spearman")
lowerCor(bfi.com,method="spearman")
print("Polychoric estimated by 2-step MLE in psych:")
lowerMat(polychoric(bfi.com)$rho)

L1 = diag(5)
for (i in 2:5) {
  for (j in 1:(i-1)) {
    x = bfi.com[,c(i,j)]
    X = t(as.matrix(x))
    L1[i,j] = esti_polychoric(X)$rho
  }
}
print("Polychoric estimated by IRLS:")
round(L1,2)
```

16 ability items scored as correct or incorrect
```{r}
ab.com <- ability[complete.cases(ability[1:5]),1:5]
print("Pearson:")
lowerCor(ab.com) #the default Pearson
print("Polychoric estimated by 2-step MLE in psych:")
lowerMat(polychoric(ab.com)$rho)

L2 = diag(5)
for (i in 2:5) {
  for (j in 1:(i-1)) {
    x = na.omit(ab.com[,c(i,j)])
    X = t(x)
    L2[i,j] = esti_polychoric(X)$rho
  }
}
print("Polychoric estimated by IRLS:")
round(L2,2)
```

The third dataset is IRLS built-in dataset.

Chinese Early Childhood Environment Rating Scale
```{r}
cec = na.omit(CECERS[,c(5,7,11,13,17)])
print("Pearson:")
lowerCor(cec)
print("Polychoric estimated by 2-step MLE in psych:")
lowerMat(polychoric(cec)$rho)

L3 = diag(5)
for (i in 2:5) {
  for (j in 1:(i-1)) {
    x = cec[,c(i,j)]
    X = t(x)
    L3[i,j] = esti_polychoric(X)$rho
  }
}
print("Polychoric estimated by IRLS:")
round(L3,2)
```

